---
title: "Reproducible Research: Storm Data Analysis"
output: 
  html_document:
    keep_md: true
---
NOAA Storm Data: Health and Economic Effects  
==============================================================

## Introduction
** Does the document have a synopsis that describes and summarizes the data analysis in less than 10 sentences? **

```{r packages}
opts_chunk$set(fig.width=8, fig.height=8,dpi=144)
require(ggplot2)
require(dplyr)
require(ggmap)
```

## obtaining data
data comes from blah
```{r loaddata,cache=TRUE}
destfile='StormData.csv.bz2'
if (! file.exists(destfile))
{
    download.file('https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2',dest=destfile,method='curl')
}
stormData<-read.csv('./StormData.csv.bz2',stringsAsFactors=FALSE)

str(stormData)
```

## Data Processing

The Event type code (EVTYPE) is important, as it will be used to categorise the data. Unfortunately, there are many errors and inconsistencies in the raw data.
```{r evtNames, dependson='loaddata', cache=TRUE}
length(unique(stormData$EVTYPE))
moose<-unique(tolower(stormData$EVTYPE))
length(moose)
moose[grep('aval',moose)]
sum(grepl('tstm',moose))
sum(grepl('thunderstorm',moose))

```

There are 87 event types that differ only in the case (upper or lower) of their lettering, 'avalanche' is missspelled at least once, and high winds may be categorised as something like 'tstm wind' or 'thunderstorm wind', with variations based on wind speeds/gust speeds. All of these variations make it difficult to compare weather events. According to the [instructions](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf) on storm data preparation, there are only 48 allowed event types, which are listed below: 

- Astronomical Low Tide 
- Avalanche 
- Blizzard 
- Coastal Flood 
- Cold/Wind Chill 
- Debris Flow 
- Dense Fog 
- Dense Smoke 
- Drought 
- Dust Devil 
- Dust Storm 
- Excessive Heat 
- Extreme Cold/Wind Chill 
- Flash Flood 
- Flood 
- Frost/Freeze 
- Funnel Cloud 
- Freezing Fog 
- Hail 
- Heat 
- Heavy Rain 
- Heavy Snow 
- High Surf 
- High Wind 
- Hurricane (Typhoon)
- Ice Storm 
- Lake-Effect Snow 
- Lakeshore Flood 
- Lightning 
- Marine Hail 
- Marine High Wind 
- Marine Strong Wind 
- Marine Thunderstorm Wind 
- Rip Current 
- Seiche 
- Sleet 
- Storm Surge/Tide 
- Strong Wind 
- Thunderstorm Wind 
- Tornado 
- Tropical Depression 
- Tropical Storm 
- Tsunami 
- Volcanic Ash 
- Waterspout 
- Wildfire 
- Winter Storm 
- Winter Weather


The raw data will need to be tidied before it can be processed effectively. First the observations (rows) of interest are selected (as this will limit the range of EVTYPE codes that need to be corrected), and then regular expressions are used to correct the event type codes to one of the 48 options listed above. Events that don't correspond to these event types will be removed, as they are not "allowed" by the NOAA and so may not have been recorded consistently (the events of these types that are included may not give an accurate representation of these storm types). 

Based on [this post](https://www.coursera.org/learn/reproducible-research/discussions/weeks/4/threads/IdtP_JHzEeaePQ71AQUtYw) in the course discussion forums, it was only after January 1996 that NOAA started recording events of all types. Tornado data was present from the beginning (1950), but other types of weather events were reported and recorded later. Data prior to January 1996 will be omitted, as analysis of this data could introduce bias due to lack of records on certain weather types.

```{r storm1996, cache=TRUE, dependson='evtNames'}
library(dplyr)
stormData$BGN_DATE<-as.Date(stormData$BGN_DATE,format='%m/%d/%Y')
storm96<-stormData %>% filter(BGN_DATE >  '1996-01-01')

```
Crop and property damage are stored strangely, with the first few significant digits stored seperately from the dollar exponent.** blah blah **

```{r storm96Interesting, cache=TRUE, dependson='storm1996'}
storm96$CropDamage<-storm96$CROPDMG
levels(as.factor(storm96$CROPDMGEXP) )

# scale crop damage
thelist<-with(storm96, CROPDMGEXP=='B')
storm96$CropDamage[thelist]<-storm96$CropDamage[thelist]*1.0e9
thelist<-with(storm96, CROPDMGEXP=='K' )
storm96$CropDamage[thelist]<-storm96$CropDamage[thelist]*1.0e3
thelist<-with(storm96, CROPDMGEXP=='M')
storm96$CropDamage[thelist]<-storm96$CropDamage[thelist]*1.0e6

# scale property damage
storm96$PropDamage<-storm96$PROPDMG
levels(as.factor(storm96$PROPDMGEXP) )

thelist<-grepl('[bB]',storm96$PROPDMGEXP)
storm96$PropDamage[thelist]<-storm96$PropDamage[thelist]*1.0e9
thelist<-grepl('[mM]',storm96$PROPDMGEXP)
storm96$PropDamage[thelist]<-storm96$PropDamage[thelist]*1.0e6
thelist<-grepl('[kK]',storm96$PROPDMGEXP)
storm96$PropDamage[thelist]<-storm96$PropDamage[thelist]*1.0e3

# slim data based on injuries, fatalities, and property/crop damage
slim96<- storm96 %>% select(BGN_DATE,EVTYPE,PropDamage,CropDamage,INJURIES,FATALITIES,LATITUDE,LONGITUDE) %>%
filter(PropDamage>0 | CropDamage >0 | INJURIES >0 | FATALITIES >0)
length(unique(slim96$EVTYPE))

```
Right now there are 222 event types. Start by casting everything to lower case, and by removing any leading whitespace

```{r evntCasting1, cache=TRUE, dependson='storm96Interesting'}
slim96$EventType<-tolower(slim96$EVTYPE)
# trimws is new? doesn't exist in older R versions
#slim96$EventType<-trimws(slim96$EventType)
slim96$EventType<-gsub('^\\s+|\\s+$','',slim96$EventType)
length(unique(slim96$EventType))
```

Reclassify/filter things
```{r cleaning, cache=TRUE, dependson='evntCasting1'}

# remove things that can't be reclassified easily
badcodes<-c("astronomical high tide","other","marine accident", "coastal storm","coastalstorm","beach erosion","glaze","mixed precip","dam break","coastal erosion" )
slim96<- slim96 %>% filter(! (EventType  %in% badcodes))

# change non thunderstorm winds to strong winds
slim96$EventType<-gsub('non[ -]tstm wind','strong wind',slim96$EventType)

#change tstm/thunderstorm wind to thunderstorm wind
change<-with(slim96, (grepl('tstm',EventType) | grepl('thunderstorm',EventType)) & grepl('wind',EventType))
slim96$EventType[change]<-'thunderstorm wind'

# change 'blowing dust' to dust devil
slim96[slim96$EventType=='blowing dust','EventType']<-'dust devil'

#change mudslide or similar to debris flow
slim96$EventType<-gsub('mud[ -]?slides?','debris flow',slim96$EventType)

# change any remaining occurences of "coastal" to coastal flood
slim96[grepl('coastal|tidal|cstl',slim96$EventType),'EventType'] <-'coastal flood'

# change (non extreme) cold to cold/windchill
slim96[grepl('cold|thermia',slim96$EventType) & (! grepl('extreme',slim96$EventType)),'EventType'] <-'cold/wind chill'

# extreme cold
slim96$EventType[grepl('extreme',slim96$EventType)] <-'extreme cold/wind chill'

# land or rock slides to debris flow
slim96$EventType[grepl('landsl|rock',slim96$EventType)] <-'debris flow'

# fog to dense fog
slim96[grepl('fog',slim96$EventType),'EventType'] <-'dense fog'

# blowing dust or whirlwind to dust devil
slim96$EventType[grepl('whirl|dust',slim96$EventType)]<-'dust devil'

# record heat to excessive heat
slim96$EventType[grepl('record heat',slim96$EventType)]<-"excessive heat"

# flash floods
slim96$EventType[grepl('flash',slim96$EventType)]<-"flash flood"

# non flash floods
slim96$EventType[grepl('high water|flood|fld',slim96$EventType) & (! grepl('flash|coastal',slim96$EventType))]<-"flood"

# frost/freeze
slim96$EventType[grepl('agric|black|road|free?z|frost',slim96$EventType) ]<-"frost/freeze"

# "landspout"
slim96$EventType[grepl('landspout',slim96$EventType)]<-"funnel cloud"

# hail
slim96$EventType[grepl('hail',slim96$EventType)]<-"hail"

# excessive heat
slim96$EventType[grepl('record',slim96$EventType)]<-"excessive heat"

# regular heat
slim96$EventType[grepl('heat|warm',slim96$EventType) & (! grepl('excessive',slim96$EventType))]<-"heat"

# rain (not gusty or freezing)
slim96$EventType[grepl('rain',slim96$EventType) & (! grepl('wind|freez',slim96$EventType))]<-"heavy rain"

# lake effect snow
slim96$EventType[grepl('lake effect snow',slim96$EventType)]<-"lake-effect snow"

# winter weather
slim96$EventType[grepl('blowing|falling|light|mix',slim96$EventType)]<-"winter weather"

# heavy snow
slim96$EventType[grepl('snow',slim96$EventType) & (! grepl('lake',slim96$EventType))]<-"heavy snow"

# high surf
slim96$EventType[grepl('surf|sea|wave|swell',slim96$EventType)]<-"high surf"

# high wind
slim96$EventType[grepl('grad|high wind',slim96$EventType)]<-"high wind"

# strong wind
slim96$EventType[grepl('gust|wind damage',slim96$EventType)]<-"strong wind"
slim96$EventType[ slim96$EventType =='wind' ]<-'strong wind'
slim96$EventType[ slim96$EventType=='winds']<-'strong wind'
slim96$EventType[ slim96$EventType== 'strong winds']<-'strong wind'

# hurricane (typhoon)
slim96$EventType[grepl('hurricane|typhoon',slim96$EventType)]<-"hurricane (typhoon)"

# rip currents (assume that drowning belongs here)
slim96$EventType[grepl('rip current|drown',slim96$EventType)]<-"rip current"

# storm surge
slim96$EventType[grepl('storm surge',slim96$EventType)]<-"storm surge/tide"

# thunderstorm winds - microbursts/downbursts are often associated with thunderstorms, so we include those here (wikipedia)
slim96$EventType[grepl('burst',slim96$EventType)]<-'thunderstorm wind'
slim96$EventType[ slim96$EventType== 'thunderstorm']<-'thunderstorm wind'

# wildfire
slim96$EventType[grepl('fire',slim96$EventType)]<-'wildfire'

#print the current list of event types
moose<-unique(slim96$EventType)
moose[order(moose)]                  

```

The following event codes are absent from the data:
 1. dust storm
 2. freezing fog
 3. lakeshore flood
 4. lightning
 5. marine hail
 6. marine high wind
 7. marine thunderstorm wind
 8. sleet


## Results
Do the figure(s) have descriptive captions (i.e. there is a description near the figure of what is happening in the figure)?
Does the analysis address the question of which types of events are most harmful to population health?
Does the analysis address the question of which types of events have the greatest economic consequences?

** axis labels and units on all plots **
What are the costliest (individual) events?
```{r summary,dependson='cleaning',cache=TRUE}
slim96$TotalDamage<-slim96$CropDamage + slim96$PropDamage
slim96 %>% select(BGN_DATE,EventType,TotalDamage,LATITUDE,LONGITUDE) %>% arrange(desc(TotalDamage)) %>% head(n=10)

```
Many of these are from the [2005 Atlantic hurricane season](https://en.wikipedia.org/wiki/2005_Atlantic_hurricane_season)
Hurricane Katrina was active during August 23-29, 2005, which accounts for 5 of 10 storms on this list. The October 24 storm (#5) corresponds to hurricane Rita, and the the 150 billion flood dated January first, 2006 is also [probably Katrina related] (https://www.ncdc.noaa.gov/billions/events).

```{r plot1, dependson='cleaning',cache=TRUE}
AggStorms<- slim96 %>% select(EventType,TotalDamage,FATALITIES,INJURIES) %>% group_by(EventType) %>% 
summarise_each(funs(sum),TotalDamageSum=TotalDamage,TotalInjuries=INJURIES,TotalFatalities=FATALITIES)
head(AggStorms)
damageStorms<-AggStorms %>% arrange(desc(TotalDamageSum)) %>% head(n=20)

g<-ggplot(data=damageStorms,aes(x=EventType,y=TotalDamageSum,fill=EventType))
g+ geom_bar(stat='identity') + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + guides(fill=FALSE)
```

```{r plot1a, dependson=c('cleaning','plot1')}
library(ggplot2)
library(dplyr)


casStorms<-AggStorms %>% mutate(casualties=TotalInjuries + TotalFatalities) %>% arrange(desc(casualties)) %>% head(n=20)

g<-ggplot(data=casStorms,aes(x=EventType,y=casualties,fill=EventType))
g+ geom_bar(stat='identity') + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + guides(fill=FALSE)
```




questions: what event types are worst for population health
panel plot fatalities vs magnitude, colour = evtype
injuries vs magnitude, colour = evetype

what storms do most damage/are most expensive
plot propdamageexp vs magnitude, colour = eventtype

where do these occur most
plot lat and long, colour = mag, on a map of US

(plots are made using [ggmap](https://journal.r-project.org/archive/2013-1/kahle-wickham.pdf) and [ggplot2](http://ggplot2.org/))


```{r mapplotsa, cache=TRUE}
#
floodnados<-slim96 %>% filter(EventType %in% c('flood','tornado')) %>% mutate(Lat=LATITUDE/100.0,Lon=-1.0*LONGITUDE/100.0)
# not 100% sure that the lat/long mapping is correct. Think that the Longitude needs that factor of -1. The LATTITUDE/LONGITUDE variables may be in degrees*100 + minutes, rather than decimal degrees*100. If so, then our coordinates might be off by up to 0.4 degrees. So there might be some inaccuracy in the coordinates used for plotting, but it shouldn't be a big deal.
mapus2<-ggmap(get_map("usa",zoom=4),extent='device')
mapus2 + stat_density2d(data=floodnados,aes(x=Lon,y=Lat,fill=..level..,alpha=..level..,colour=EventType),bins = 6, geom = "polygon") + facet_wrap(~EventType,ncol=1) + theme(legend.position='none') + theme()


```

